---
title: "ShoalsData_2011to2024_PilotAnalysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

10/14/24

Uploading the data
```{r}

setwd("C:/Users/jessv/OneDrive/Desktop/BIO 809")
shoals <-read.csv("2024_Shoals_SealSurveys.csv")
```

Clean up the data to survey counts per year with standard deviation, mean, etc. 

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)

shoals_avg <- shoals %>% 
  group_by(Year) %>% 
  summarize(Pv= mean(TOTAL.PV), Hg= mean(TOTAL.HG))

shoals_sd <- shoals %>% 
  group_by(Year) %>% 
  summarize(Pv=sd(TOTAL.PV), Hg=sd(TOTAL.HG))

shoals_avlong <- shoals_avg %>% 
  pivot_longer(names_to= "Species", values_to = 'meancount', cols=Pv:Hg) 

shoals_sdlong <- shoals_sd %>% 
  pivot_longer(names_to='Species', values_to = 'sdcount', cols=Pv:Hg)

shoals_joined <- inner_join(shoals_avlong, shoals_sdlong)

ggplot(data=shoals_clean, aes(x=Year, y=AvgPv:AvgUnk)) +
    geom_bar(stat='identity', position=position_dodge()) 


ggplot(data=shoals_joined, aes(x=Year, y=meancount, color=Species)) +
  geom_point() + geom_smooth(method = 'glm', method.args=('poisson')) +
  ylab("Average Count") +
  xlab("Year") +
  ggtitle("Average Count of Seals per Year") +
  scale_color_discrete(labels = c("Hg" = "Gray Seal", "Pv" = "Harbor Seal"))

ggplot(data=shoals, aes(x=TOTAL.HG)) +
  geom_histogram(color= 'darkblue', fill="blue", bins=15) + 
  geom_vline(aes(xintercept=mean(TOTAL.HG)), color="navyblue", linetype="dashed", size=1) +
  xlab("Gray Seal Count per Survey") +
  ylab("Count") + 
  ggtitle("Gray seal count histogram 2011-2024")

ggplot(data=shoals, aes(x=TOTAL.PV)) +
  geom_histogram(color= 'green4', fill="limegreen", bins=15) + 
  geom_vline(aes(xintercept=mean(TOTAL.PV)), color="darkgreen", linetype="dashed", size=1) +
  xlab("Harbor Seal Count per Survey") +
  ylab("Count") + 
  ggtitle("Harbor seal count histogram 2011-2024")

hist(shoals$TOTAL.HG)
hist(shoals$TOTAL.PV)

#linear model output
mymod <- lm(TOTAL.HG ~ Year, data = shoals)
summary(mymod)
#according to linear model, for each year the count of Hg increases by 14.886
pvmod <- lm(TOTAL.PV ~ Year, data=shoals)
summary(pvmod) #for each year there is a 7.395 decrease in Pv

hgmodglm <- glm(TOTAL.HG ~ Year, data = shoals, family= 'poisson')
summary(hgmodglm)

#the rate at which the count is changing because this is exponentiatedd
#Get rid of log term 
exp(8.358e-02)
#result: 1.0872
#increasing at a rate of 8.7% per year 

pvmodglm <- glm(TOTAL.PV ~ Year, data=shoals, family = 'poisson')
summary(pvmodglm)
exp(-0.0326890)
#result: 0.9678
1-.9678
#Population is decreasing at a rate of 3.2% per year 

speciesmod <- glm(TOTAL.HG ~ Year + TOTAL.PV, data=shoals, family= 'poisson')
summary(speciesmod)

#Year
exp(1.031e-01)
#1.108602
#when the effect of pv is held constant, hg are increasing at a rate of 10.87% per year
#p value: <2e-16

#Pv
exp(2.329e-03)
#1.002332
#When the effect of year is held constant, Hg increase 0.23% with every one unit increase in Pv 
# p value: <2e-16 

pvmod <- glm(TOTAL.PV ~ Year + TOTAL.HG, data=shoals, family='poisson')
summary(pvmod)

#Year 
exp(-7.090e-02)
#0.931555
1-0.931555
#0.068445
#With effect of hg held constant, Pv decrease at rate of 6.8% per year 
#p=<2e-16

#hg
exp( 2.258e-03)
#1.002261
#When effect of year is held constant, Pv increase at rate of 0.226% with every one unit increase in Hg 
```


Making a prettier plot using old script
```{r}
#need to remove 2014 

shoals_joined <- shoals_joined[-c(7,8),]

#plotting
ggplot(data=shoals_joined, aes(x=as.factor(Year), y=meancount, fill=Species)) + 
  geom_bar(stat="identity", position=position_dodge(), width=0.8)+
  scale_fill_manual(values=c("slategray", "slategray1", "slategray2")) + 
  theme_minimal() + 
  ggtitle("Average Number of Seals Per Year")+
  ylab("Number of Seals") + 
  geom_errorbar(aes(ymin=meancount-sdcount, ymax=meancount+sdcount), width=.2,
                position=position_dodge(.9)) +
  scale_x_discrete(name ="Year", breaks=c(2011, 2012, 2013, 2015,
                                          2016, 2017, 2018, 2019, 
                                          2022, 2023, 2024), 
                   labels=c("2011","2012","2013", "2015", 
                            "2016", "2017", "2018", "2019", 
                            "2022", "2023", "2024"), drop=FALSE)
  
  xlim("2011","2012","2013", "2015", "2016", "2017", "2018", "2019", 
       "2022", "2023", "2024")

  scale_x_discrete(name ="Year", breaks=c(2011, 2012, 2013, 2015,
                                          2016, 2017, 2018, 2019, 
                                          2022, 2023, 2024), 
                   labels=c("2011","2012","2013", "2015", 
                            "2016", "2017", "2018", "2019", 
                            "2022", "2023", "2024"))

ggplot(data=shoals_joined, aes(x=as.factor(Year), y=meancount, color=Species, 
                               group=Species)) +
  geom_point() + geom_line()

```

Okay, attempting to merge SST data with shoals data.

```{r}
Temp <- read.csv("6.2024to8.2024_WMS_SST_AT.csv")

Temp <- Temp %>% 
  mutate(Temp, station='BO1')

ggplot(data=Temp, aes(x=Date, y=BO1.1M.Daily.Average.SST, group=station)) +
  geom_line() 

shoals24 <- shoals[-c(1:256),]

shoals24clean <- shoals24 %>% 
  select(Date, TOTAL.PV, TOTAL.HG) %>% 
  pivot_longer(names_to= "Species", values_to = 'count', cols=TOTAL.PV:TOTAL.HG)

tempjoin <- inner_join(Temp, shoals24clean)

coeff <- .04

ggplot(tempjoin, aes(x=Date)) +
  geom_point(aes(y=count, group=Species, color=Species), size=1.5) + geom_line(aes(y=count, group=Species, color=Species), size=.8) +
  geom_line(aes(y=BO1.1M.Daily.Average.SST/coeff, group=station, color=station), size=.8) +
  scale_y_continuous(name="Minimum Count Estimate",
                     sec.axis=sec_axis(~.*coeff, name="Temperature (C)")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(axis.title.y= element_text(color="blue"),
        axis.title.y.right=element_text(color="red")) +
  scale_color_discrete(labels = c("TOTAL.HG" = "Gray Seal", "TOTAL.PV" = "Harbor Seal", 
                                  "BO1" = "SST")) +
  labs(color = "Key") +
  ggtitle("Minimum count estimates and SST during 2024 survey season") 

```

Making boxplots

```{r}
shoalsbox <- shoals %>% 
  select(Year, TOTAL.PV, TOTAL.HG) %>% 
  pivot_longer(names_to= "Species", values_to = 'count', cols=TOTAL.PV:TOTAL.HG) 

shoalsboxno2014 <- shoalsbox[-c(151:164),]
  

ggplot(shoalsboxno2014, aes(y=count, fill=Species)) +
  geom_boxplot() +
  facet_wrap(~Year)

ggplot(shoalsboxno2014, aes(x=as.factor(Year), y=count, fill=Species)) +
  geom_boxplot() +
  facet_wrap(~Species)

ggplot(shoalsboxno2014, aes(x=as.factor(Year), y=count, fill=Species)) +
  geom_boxplot()

```

Trying to run a model with count data and temperature data. 

```{r}

shoals24wide <- shoals24 %>% 
  select(Date, TOTAL.PV, TOTAL.HG) 
tempjoin2 <- inner_join(Temp, shoals24wide)
tempmod <- glm(TOTAL.HG ~ BO1.1M.Daily.Average.SST, data = tempjoin2, family= 'poisson')
summary(tempmod)
exp(-0.076168)
#0.9266
1-0.9266
#For every one unit increase in SST, hg count declines by 7.3%

tempmod2 <- glm(TOTAL.HG ~ BO1.1M.Daily.Average.SST + TOTAL.PV, data = tempjoin2, family= 'poisson')
summary(tempmod2)
AIC(tempmod, tempmod2)

#SST: -0.0152134
exp(-0.0152134)
1-.9849017
#For every one unit increase in SST, Hg count declines by 1.5% (not sig)

#PV: 0.0045475  
exp(0.0045475)
#1.004558
#With SST held constant, for every additional harbor seal, gray seals increase by .45%
#does that make sense

tempmod3 <- glm(TOTAL.PV ~ BO1.1M.Daily.Average.SST + TOTAL.HG, data = tempjoin2, family= 'poisson')
summary(tempmod3)

#sst
exp(-0.0082244)
#0.9918093
1-0.9918093
#0.0081907
#p=0.444    

#hg
exp( 0.0030177)
#1.003022
#with sst held constant, for every additional gray seal, pv increase by 0.3%
```


Trying to do something now with ledge per seal...wish me luck.


```{r}

ledgedata <- shoals %>% 
  select(Year, Pv.....1S, PV.2N, PV.3S, PV.3N, PV.4E, PV.4NE, PV.4NW, PV.6, Pv...1N, Pv...2S, Pv.2E, 
         HG.2E, HG.2N, HG.3S, HG.3N, HG.4E, HG.4NE, HG.4NW, HG.6, Hg...1S, Hg...1N, Hg.2S)


```

Making a figure with count per month combined by all years

```{r}
#First I need to make a new column with month extracted from date

# Load the lubridate package
library(lubridate)
# Create a sample date
#date <- ymd("2023-11-30")
# Extract the month using lubridate's month() function
#month <- month(date)
# Print the result
#print(month

shoals$month <- format(as.Date(shoals$Date, format="%m/%d/%Y"),label=TRUE, "%m")

shoalscpmav <- shoals %>% 
  group_by(month) %>% 
  summarize(Pv = mean(TOTAL.PV), Hg=mean(TOTAL.HG))

shoalscpmsd <- shoals %>% 
  group_by(month) %>% 
  summarize(Pv = sd(TOTAL.PV), Hg=sd(TOTAL.HG))

shoalscpmavlong <- shoalscpmav %>% 
  pivot_longer(names_to="Species", values_to="meancount", cols=Pv:Hg)

shoalscpmsdlong <- shoalscpmsd %>% 
  pivot_longer(names_to="Species", values_to="sdcount", cols=Pv:Hg)

cpmjoined <- inner_join(shoalscpmavlong, shoalscpmsdlong)

ggplot(data=cpmjoined, aes(x=month, y=meancount, fill=Species)) +
  geom_bar(stat="identity", position=position_dodge()) +
  geom_errorbar(aes(ymin=meancount-sdcount, ymax=meancount+sdcount), width=.2,
                 position=position_dodge(.9)) + 
  scale_fill_manual(values=c("slategray", "slategray1", "slategray2")) +
  theme_minimal() +
  ggtitle("Average number of seals per month from 2011-2024") +
  ylab("Number of seals") +
  xlab("Month") +
  scale_x_discrete(labels=c("May", "June", "July", "August")) 
```

Table with number of surveys 
```{r}
shoalscpm <- shoals %>% 
  group_by(month) %>% 
  summarize(count= n(), Pv = mean(TOTAL.PV), sdPv=sd(TOTAL.PV), Hg=mean(TOTAL.HG),sdHg=sd(TOTAL.HG))

library(knitr)
library(kableExtra)
kable(shoalscpm, col.names=c( "Month", "Number of surveys per month", "Average Harbor Seal Count", "Standard Deviation of Harbor Seal Count", "Average Gray Seal Count", "Standard Deviation of Gray Seal Count")) %>% 
  kable_classic(full_width=F, html_font="Cambria")
```

